Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-14 programlogic.tex
4.3 revised

We place the authorative ownership $\mathcal{A}\textsf{PTableWalk}$ of the ghost-map in a per-address-space invariant
($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}) whose ownership, when combined with the
fraction stored in the virtual points-to, permits changes in the virtual-to-physical translations.
By itself, it allows changes to the page tables as long as they do not change the overall mappings from virtual
to physical addresses.
\todo[inline]{explain $\Theta$, $m$, and figure out where $\delta$ is bound in Figure \ref{fig:peraspaceinvariant}}

\

@@

  % Colin removed this remark because it's skwtching a *3rd* alternative design for virtual points-to,
  % which could be useful for some readers, but will most likely just confuse
% \begin{remark}[Exchanging Page-Table-Walk-Summarization Mappings (Tokens) under Mutation of Page Tables ]
%   $\mathcal{A}\textsf{PTableWalk}$'s fragmental-ownership represents an individual key-value pair (per-virtual-address) which can be handled as a \textit{look-up} resource for an individual virtual address.
%   To build more intuition on the convenience of this protocol, we can think of the use-case \textit{unmapping-pages}. Normally, in the existence of shared page-tables, we need to first collect all the fragmental map pairs, e.g. $\sumwalkabs\vaddr\qfrac\paddr$, then, after making sure of non-existence of \textit{sharing}, we can \textit{unmap} the page. In this case, it would be ok to put $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ and $ \sumwalkabs\vaddr\qfrac\paddr$ side-by-side inside the virtual-pointsto definition in Figure \ref{fig:virtualpointstosharing}. However, this pattern of verifying is not only inconvenient due to requirement of extra bookkeeping, but also violates our soundness argument based on considering  mutation on the shared page-tables -- due to coalescing of page tables etc. --. In other words, physical $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ sitting aside to  $ \sumwalkabs\vaddr\qfrac\paddr$ would be made invalid by any concurrent update on the physical page-table. So there should only be one physical memory resource inside the virtual pointsto definition: from physical page address to the value stored ($ \paddr \mapsto_{p} \vpage$).
% \end{remark}

%\subsection{Address-Space Abstraction}
%\label{sec:aspacemodal}

Relations amongst these logical constructions, for the sake making it clearly understood, outlined pictorially  in Figure \ref{fig:logicaladdrspace}:
\begin{itemize}
\item a global address space invariant (solid round square labelled as $\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:logicaladdrspace}) asserting the authorative ownership (i.e. update capability), $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$, of the ghost map -- shown as solid-round-head arrow to $\mathcal{A}\textsf{PTableWalk}$ Figure \ref{fig:logicaladdrspace}
\item per-virtual-pointsto relation ((cutted round square labelled as \textsf{Virtual PointsTo} in Figure \ref{fig:peraspaceinvariant})) asserting the fragmental ownership of the ghost map, $\sumwalkabs\vaddr\qfrac\paddr$ \textit{represents} the fragmental-ownership (cutted-diamond-head arrow) of $\mathcal{A}\textsf{PTableWalk}$ 
\item there exists only one persistent authorative ownership ($\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$) for all fragmental ghost page-table-walk mapping ($\sumwalkabs\vaddr\qfrac\paddr$) shown as double-ended solid arrow 1 to N imposing the update protocol
\end{itemize}

\begin{figure}
   \includegraphics[width=0.75\columnwidth]{logical_addr_space.png}
  \caption{Logical Resources of Address-Space Abstraction}
  \label{fig:logicaladdrspace}
  \end{figure}

>>>>

We place the authorative ownership $\mathcal{A}\textsf{PTableWalk}$ of the ghost-map in a per-address-space invariant
($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}) whose ownership, when combined with the
fraction stored in the virtual points-to, permits changes in the virtual-to-physical translations.
By itself, it allows changes to the page tables as long as they do not change the overall mappings from virtual
to physical addresses.

@@

Figure \ref{fig:peraspaceinvariant} describes the resources for each address space, which are collected
for the full address space, but separated from virtual points-to assertions.
$\Theta$ is the logical map from virtual addresses to the physical addresses they should translate to, 
for the \emph{current} address space, corresponding to the currently installed page tables (the ones
indicated by \lstinline|cr3|). In Figure \ref{fig:virtualpointstosharing}, the existentially-quantified $\delta$
is the ghost name for the authoritative copy of this map (again, for the current address space).
$m$ is the (full) logical map from the various page table bases to ghost names for the 
per-address-space mappings like $\Theta$.
It is the full version of the map named by the ghost name $\delta{}s$ in
Figure \ref{fig:virtualpointstosharing}; there is only one such map in the system.
This is why Figure \ref{fig:virtualpointstosharing} asserts there is some ghost name in that (global) map
for the current address space.

$\mathcal{I}\textsf{ASpace}(\Theta,m)$ is then the resources for the current address space:
It ensures that the current address space is valid (it has an entry in $m$, whose corresponding
authoritative map matches $\Theta$), and asserts ownership of the appropriate fractional ownership
of the physical page table walk for each virtual-to-physical translation in $\Theta$ (for the current address space).

Figure \ref{fig:logicaladdrspace} depicts the relationships between these pieces, for a single
address space.
The invariant for the currently active address space (solid round square labelled as 
$\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:logicaladdrspace}) asserts authorative ownership (i.e. 
update capability), $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$, of the ghost map $\Theta$
describing virtual-to-physical address translations
-- shown as a solid-round-head arrow to $\mathcal{A}\textsf{PTableWalk}$.
Each virtual points-to assertion (dashed round square labelled as \textsf{Virtual PointsTo}
 in Figure \ref{fig:peraspaceinvariant})
asserts fragmental ownership of the ghost map $\Theta$, $\sumwalkabs\vaddr\qfrac\paddr$,
shown via the dashed-diamond-head arrow to $\mathcal{A}\textsf{PTableWalk}$. Specifically, it asserts that
the virtual address the virtual points-to assertion concerns corresponds to some physical address.
There exists only one persistent authorative ownership 
($\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$) which collects the fragmental 
ghost page-table-walk mapping ($\sumwalkabs\vaddr\qfrac\paddr$),
 shown as double-ended solid arrow 1 to N imposing the update protocol.

\begin{figure}
   \includegraphics[width=0.75\columnwidth]{logical_addr_space.png}
  \caption{Logical Resources of Address-Space Abstraction (1 Address Space)}
  \label{fig:logicaladdrspace}
  \end{figure}

\paragraph{From A Single Address Space to Many}
In principle the resources from each individual address space should be collected into a single
shared invariant, and for each memory access,
a fragment of this corresponding to $\mathcal{I}\textsf{ASpace}(\Theta,m)$
for the current address space should be extracted from this global resource, used to prove correctness of the
memory access, and put back.
In this paper we focus on the middle section, explicity identifying resources for each
address space. The reason for this is that in practice, this global resource is also
the place where kernel-specific assumptions, such as guaranteeing that certain virtual address range
was mapped in all address spaces, would be enforced. This paper focuses on the
reasoning principles behind the virtual memory access, and we leave the use of this within a larger
kernel to future work.
